---
title: "Getting started with zufuzz"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with zufuzz}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(zufuzz)
```

Fuzzing is mechanical: generate an input, run the code, see whether something
broke, repeat. What makes it useful rather than merely random is *coverage
guidance* — the engine keeps inputs that reached somewhere new, so it builds
its way into code a random generator would never reach.

This vignette is about the parts you have to get right for that to pay off.

## The oracle: what counts as a bug

A fuzzer needs a rule for "this input broke something". zufuzz's rule is
simple and deliberately strict:

> Any condition of class `error` that escapes `test_one_input` is a finding.

Which puts the weight on how you handle input your function is *allowed* to
reject. Most parsers reject bad input by design, and if every rejection were a
finding you would get thousands of them and learn nothing.

The wrong way:

```r
test_one_input <- function(data) {
  tryCatch(parse_thing(rawToChar(data)), error = function(e) NULL)
}
```

That catches the defects too. Every segfault-adjacent bug, every
`subscript out of bounds`, every corrupted-state error — all silently
swallowed. The fuzzer will run happily for hours and report nothing.

The right way is to name the class, and wrap only the call that may raise it:

```r
test_one_input <- function(data) {
  text <- fuzzed_data_provider(data)$consume_string()

  parsed <- tryCatch(
    parse_thing(text),
    thing_parse_error = function(e) NULL   # documented rejection
  )

  if (!is.null(parsed)) {
    stopifnot(identical(parse_thing(serialize_thing(parsed)), parsed))
  }
}
```

Now a `thing_parse_error` is uninteresting and anything else is a finding —
including a `thing_parse_error` raised from the *round trip*, which would be a
real bug.

If your package signals plain `stop("...")` without a condition class, that is
worth fixing before fuzzing it: without a class there is no way to separate
"rejected as designed" from "broke".

`fuzz_function()` enforces this — it refuses `expect = "error"` outright.

## What gets instrumented, and what does not

```{r}
classify <- function(x) {
  if (x > 0) "positive" else "negative"
}
instrument("classify")

classify(1)
instrumentation_report()
```

The rewrite puts a probe at function entry, at each statement of a block, on
both outcomes of every `if`, and at loop-body entry. It preserves value,
visibility, laziness, evaluation order, error propagation, `return`/`break`/
`next`, and `on.exit`.

There are holes, and knowing them is the difference between "my coverage is
low" and "my coverage is low *because*":

* **Nested function literals are not instrumented.** Rewriting inside
  `lapply(xs, function(x) ...)` would change what `substitute()` sees of the
  argument. They are reported instead, so closure-heavy code tells you why its
  coverage looks thin.
* **Call arguments are left alone**, for the same reason.
* **Conditions are entered only to find comparisons**, never to place probes —
  evaluating a condition twice would change the program.
* **S4, R6 and reference classes** are out of scope for this release.
* **Aliases captured before instrumentation** still point at the original.
  `instrumentation_report()` lists them.

```{r}
uninstrument()
```

## Three different claims about reproducibility

These get conflated constantly, and they are not the same claim.

| | means |
| --- | --- |
| **input reproduction** | the artifact holds exactly the bytes that were given |
| **finding reproduction** | a fresh, uninstrumented process fails the same way |
| **campaign determinism** | same engine, flags, seed and corpus → same decisions |

`replay()` answers the middle one, and it is the one that matters when you
file a bug:

```r
replay("harness.R", "crash-3f2a...")
```

It runs the harness in a new process with instrumentation off. A finding that
only reproduces against rewritten code is a finding about zufuzz, not about
your package. A finding that depends on what ran *before* it comes back not
confirmed, which is also worth knowing before you go hunting.

## Choosing an engine

```{r}
engines()
```

`"none"` always works: it runs the inputs you name, once each, and returns.
That is what `replay()` uses and what makes Windows usable.

`"afl"` needs `afl-fuzz` installed. Two things catch people out, and zufuzz
warns about both rather than letting them look like success:

* A harness that hardcodes `engine = "none"` will read its inputs and exit
  without ever speaking the fork-server protocol. Use `engine = "auto"`.
* A harness that instruments nothing gives AFL an empty coverage map, and AFL
  aborts with "no instrumentation detected". The campaign never starts.

## Structured input

```{r}
spec <- r_object(types = c("integer", "character"), max_len = 4)
xs <- draw(spec, n = 3, seed = 42)
str(xs)
```

Every object is a pure function of its bytes, so it has an exact on-disk
form. Keep the interesting ones as corpus seeds with `as_seed()`, and turn a
crash artifact back into the object that caused it with `object_from()`.

The `nasty` validity level — the default — includes `NA` of every type, `NaN`,
`±Inf`, `-0`, zero-length vectors, duplicate and empty names, and `latin1`
encoding marks. All of that is legal R that most code never sees in testing. A
crash on any of it is a defect. `strict` restricts generation to what base
constructors produce.

## Native code and sanitizers

zufuzz never owns a sanitizer; it arranges for one to be present.

The simplest configuration is a sanitized R build (`rocker/r-devel-san`,
`wch1/r-debug`) running as an AFL worker. Nothing is preloaded, because
zufuzz's own compiled code references no sanitizer symbol at all.

```sh
ASAN_OPTIONS=detect_leaks=0:allocator_may_return_null=1:abort_on_error=1 \
  afl-fuzz -i corpus -o out -- Rscript harness.R
```

`detect_leaks=0` because R leaves allocations for the OS at exit and the
reports are noise; `allocator_may_return_null=1` so R's own out-of-memory path
is reached rather than an abort.

For the R-specific bug class no sanitizer sees — a missing `PROTECT` — use
`fuzz(gc_torture = TRUE)`. Expect a large slowdown, and pair it with ASan
rather than choosing between them.

## Windows

Everything except a campaign. You can instrument, generate objects, replay an
artifact a colleague found, minimize it, and report coverage. No fuzzing
engine runs under Rtools, so campaigns want WSL or a container.

```{r}
engine_available("none")
```
